# Pseudobulk pt 2
Pseudobulk analysis of resolving RPRA vs non-resolving RPRA samples from [Bailey et al. bioRxiv 2023](https://www.biorxiv.org/content/10.1101/2023.07.30.551145v1.full#F1).

In this notebook, I import the pseudobulk counts (gene sums) and number of cells for each sample generated in the previous notebook and use this information to:
- filter out: 
  - samples containing < 10 cells, 
  - cell types containing only one status,
  - cell types with only one sample for a particular status
- plot:
  - number of cells per sample in each cell type
  - number of genes detected in each sample
  - number of unique counts per library
  - heatmap of library distances
  - local and parametric fit
  - PCA
  - MA plot
- output (as csv):
  - DEGs
  - DEGs from MA plot (padj < 0.05)
  
Output can be found here:
`../../../data/22deg-analysis/bailey_resolv_vs_nonresolv_v2/pseudobulk/ouput`
  
See next notebook for further analysis.

## Import packages

```{r}
library(DESeq2)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(R.utils)
library(vsn)
```

```{r}
plot_lib_sizes <- function(data) {
    df <- data.frame(lib.size = sort(colSums(data)))
        ggplot(df, aes(x = reorder(rownames(df), lib.size), y = lib.size)) +
        geom_col() +
        ggtitle("Sorted library sizes") +
        theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1)) +
        xlab("")
}
```

```{r}
DATA_DIR <- "../../../data/22deg-analysis/bailey_resolv_vs_nonresolv_v2/pseudobulk/input"
OUT_DIR <- "../../../data/22deg-analysis/bailey_resolv_vs_nonresolv_v2/pseudobulk/output"
```


```{r}
process_cell_type <- function(cell_type) {
  print(sprintf("Processing %s", cell_type))
  data <- read.table(sprintf("%s/%s.txt", DATA_DIR, cell_type), header = TRUE, check.names = FALSE)
  coldata <- read.csv(sprintf("%s/%s-meta.csv", DATA_DIR, cell_type), check.names = FALSE)
  
  if (length(unique(coldata$Status)) < 2) {
    print("Only one Status, skipping")
    return()
  }

  if (count(coldata$Status == 'RPRA resolving') < 2) {
    print("Only one RPRA resolving sample, skipping")
    return()
  }

  if (count(coldata$Status == 'RPRA non-resolving') < 2) {
    print("Only one RPRA non-resolving sample, skipping")
    return()
  }

  dir <- sprintf("%s/%s", OUT_DIR, cell_type)
  R.utils::mkdirs(dir)
  write.csv(
    coldata,
    sprintf("%s/meta.csv", dir)
  )
  
  counts <- as.matrix(data[, coldata[["sample"]]])
  dds <- DESeqDataSetFromMatrix(round(counts), coldata, ~ Status) 

  pdf(sprintf("%s/lib-sizes.pdf", dir), width = 6, height = 4)
  plot(plot_lib_sizes(counts))
  dev.off()

  # Library complexity
  # TODO: explore other measures of library complexity, like dispersion, skew??
  d <- data.frame(
    n_genes = colSums(counts(dds) > 0),
    uniq_counts = apply(counts(dds), 2, function(x) {return(length(unique(x)))})
  )

  # Plot number of genes detected in each sample
  pdf(sprintf("%s/n_genes.pdf", dir), width = 6, height = 4)
  plot(ggplot(d, aes(x = reorder(rownames(d), n_genes), y = n_genes)) +
    geom_col() +
    theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1)) +
    xlab(""))
  dev.off()

  # Number of unique counts per library
  pdf(sprintf("%s/n_unique_counts.pdf", dir), width = 6, height = 4)
  plot(ggplot(d, aes(x = reorder(rownames(d), uniq_counts), y = uniq_counts)) +
    geom_col() +
    theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1)) +
    xlab(""))
  dev.off()
  
  # needed for GSVA
  transformed <- assay(vst(dds))
  write.table(transformed, sprintf('%s/%s-transformed.tsv', DATA_DIR, cell_type), sep="\t", quote=FALSE)

  # Plot heatmap of library similarity
  sampleDists <- dist(t(assay(vst(dds))))

  pdf(sprintf("%s/lib-dist.pdf", dir), width = 8, height = 8)
  pheatmap(
    as.matrix(sampleDists),
    cluster_cols = TRUE,
    cluster_rows = TRUE,
    show_colnames = FALSE,
    clustering_distance_rows = sampleDists,
    clustering_distance_cols = sampleDists
  )
  dev.off()

  dds <- DESeq(dds, fitType = "parametric")
  pdf(sprintf("%s/disp-parametric.pdf", dir), width = 6, height = 4)
  plotDispEsts(dds)
  dev.off()

  dds <- DESeq(dds, fitType = "local")
  pdf(sprintf("%s/disp-local.pdf", dir), width = 6, height = 4)
  plotDispEsts(dds)
  dev.off()

  pdf(sprintf("%s/pca.pdf", dir), width = 6, height = 4)
  d <- plotPCA(
    vst(dds, blind = FALSE),
    ntop = 2000,
    intgroup = c("Status"),
    returnData = TRUE
  )
  percentVar <- round(100 * attr(d, "percentVar"))
  plot_title <- ifelse(cell_type == "moam-1", "MoAM-1", cell_type)


# code for Figure 5b
  plot(ggplot(d, aes(
    x = PC1,
    y = PC2,
    color = Status,
  )) +
    geom_point(size = 5) +
    scale_color_manual(values = c("RPRA non-resolving" = rgb(0.90196078, 0.29411765, 0.20784314), 
                                  "RPRA resolving" = rgb(0.30196078, 0.73333333, 0.83529412)
                                  )) +
    xlab(paste0("PC1: ", percentVar[1], "% variance")) +
    ylab(paste0("PC2: ", percentVar[2], "% variance")) +
    coord_fixed() +
    guides(color = guide_legend(order = 1), shape = guide_legend(order = 2)) +
    ggtitle(plot_title) +
    theme_bw() +
      theme(
      text = element_text(size = 15),
      plot.title = element_text(hjust = 0.5),     # Center the plot title
      legend.title = element_text(size = 15),     # Set legend title font size
      legend.text = element_text(size = 15),      # Set legend text font size
      axis.title = element_text(size = 15),       # Set axis title font size
      axis.text = element_text(size = 15, color = "black"),         # Set axis text font size
      panel.grid.major = element_blank(),         # Remove major gridlines
      panel.grid.minor = element_blank()          # Remove minor gridlines
    )
  )
  dev.off()

  degs <- as.data.frame(results(dds, contrast = c("Status", "RPRA resolving", "RPRA non-resolving"), alpha = 0.05))
  degs <- degs[!is.na(degs$padj), ]
  
  write.csv(
  degs[order(degs$log2FoldChange, decreasing = TRUE), ],
  sprintf("%s/gsea_degs.csv", dir)
  )    
  
  degs <- degs[(degs$baseMean > 100), ]
  degs$sign <- ""
  degs$sign[(degs$log2FoldChange < 0) & (degs$padj < 0.05)] <- "Downregulated"
  degs$sign[(degs$log2FoldChange > 0) & (degs$padj < 0.05)] <- "Upregulated"
  degs <- degs[order(degs$sign),]

  pdf(sprintf("%s/ma.pdf", dir), width = 8, height = 6)
  max_x = max(degs$baseMean)
  plot(ggplot(degs) +
    geom_point(aes(x = baseMean, y = log2FoldChange, col = sign)) +
    scale_x_log10() +
    scale_color_manual(values = c("gray", "blue", "red")) +
    geom_text(x = 1e+05, y = 5, label = sprintf("Up in RPRA resolving: %d", sum(degs$sign == "Upregulated")), hjust = 1) +
    geom_text(x = 1e+05, y = -6, label = sprintf("Up in RPRA non-resolving: %d", sum(degs$sign == "Downregulated")), hjust = 1) +
    ggtitle("RPRA resolving vs RPRA non-resolving"))
  dev.off()

  write.csv(
    degs[order(degs$log2FoldChange, decreasing = TRUE), ],
    sprintf("%s/degs.csv", dir)
  )

  sign.only <- degs[(degs$padj < 0.05), ]
  write.csv(
    sign.only[order(sign.only$log2FoldChange, decreasing = TRUE), ],
    sprintf("%s/degs-sign.csv", dir)
  )
  print("done")
}
```

```{r}
cell_types <- list.files(DATA_DIR, pattern = "*.csv")
for (f in cell_types) {
  coldata <- read.csv(sprintf("%s/%s", DATA_DIR, f), check.names = FALSE)
  # Require samples with more than 10 cells
  coldata <- coldata[coldata$n_cells > 10, ]
  cell_type <- sub("-meta.csv", "", f)
  # Require at least 1 samples for cell type
  if (nrow(coldata) > 1) {
    process_cell_type(cell_type)
  }
}
```

### Next Steps
The next notebook will summarize the data generated in this notebook. More specifically, it will:
- Plot & return number of DEGs per cluster
- Plot & return number of filtered DEGs for each cluster
